A simplified dynamic bioenergetic model for coral-\emph{Symbiodinium} symbioses and coral bleaching as an alternate stable state

### 2016-11-01 JTB_ms/JTB_ms.tex
complete draft

\usepackage[T1]{fontenc}
\usepackage{lmodern}
\usepackage{amssymb,amsmath}
\usepackage{ifxetex,ifluatex}
\usepackage{fixltx2e} % provides \textsubscript
% use upquote if available, for straight quotes in verbatim environments
\IfFileExists{upquote.sty}{\usepackage{upquote}}{}
\ifnum 0\ifxetex 1\fi\ifluatex 1\fi=0 % if pdftex
  \usepackage[utf8]{inputenc}
\else % if luatex or xelatex
  \usepackage{fontspec}
  \ifxetex
    \usepackage{xltxtra,xunicode}
  \fi
  \defaultfontfeatures{Mapping=tex-text,Scale=MatchLowercase}
  \newcommand{\euro}{€}
\fi
% use microtype if available
\IfFileExists{microtype.sty}{\usepackage{microtype}}{}
\usepackage{longtable}
\usepackage{graphicx}
% We will generate all images so they have a width \maxwidth. This means
% that they will get their normal width if they fit onto the page, but
% are scaled down if they would overflow the margins.
\makeatletter
\def\maxwidth{\ifdim\Gin@nat@width>\linewidth\linewidth
\else\Gin@nat@width\fi}
\makeatother
\let\Oldincludegraphics\includegraphics
\renewcommand{\includegraphics}[1]{\Oldincludegraphics[width=\maxwidth]{#1}}
\ifxetex
  \usepackage[setpagesize=false, % page size defined by xetex
              unicode=false, % unicode breaks when used with xetex
              xetex]{hyperref}
\else
  \usepackage[unicode=true]{hyperref}
\fi
\hypersetup{breaklinks=true,
            bookmarks=true,
            pdfauthor={},
            pdftitle={A simplified dynamic bioenergetic model for coral-algal symbiosis and coral bleaching},
            colorlinks=true,
            urlcolor=blue,
            linkcolor=magenta,
            pdfborder={0 0 0}}
\urlstyle{same}  % don't use monospace font for urls
\setlength{\parindent}{0pt}
\setlength{\parskip}{6pt plus 2pt minus 1pt}
\setlength{\emergencystretch}{3em}  % prevent overfull lines
\setcounter{secnumdepth}{0}
% Pandoc toggle for numbering sections (defaults to be off)
\setcounter{secnumdepth}{0}
% Pandoc header

@@

  \title{A simplified dynamic bioenergetic model for coral-algal symbiosis and
coral bleaching}
    \author[University of Hawaii]{Ross Cunning\corref{c1}}
   \ead{ross.cunning@gmail.com} 
   \cortext[c1]{Corresponding Author}
    \author[University of California]{Erik B. Muller}

@@

  \begin{abstract}
  Coral reef ecosystems owe their ecological success--and vulnerability to
  climate change--to the symbiotic metabolism of corals and
  \emph{Symbiodinium} spp. The urgency to understand and predict the
  stability and breakdown of these symbioses (i.e., coral `bleaching')
  calls for the development and application of theoretical tools. While
  the biological complexity of the symbiosis is well-suited to dynamical
  modeling, the theoretical complexity of certain approaches makes them
  inaccessible to many biologists, thus limiting their potential
  application and utility. Here, we develop a simplified dynamic
  bioenergetic model of coral-algal symbiosis that demonstrates realistic
  steady-state patterns in coral growth and symbiont abundance across
  gradients of light, nutrients, and feeding. Furthermore, through a
  mechanistic treatment of photo-oxidative stress, the model displays
  dynamic responses of bleaching and recovery, and suggests that these
  phenomena represent transitions between alternative stable states. A
  suite of complex observed behaviors reproduced by the model suggests it
  captures many important attributes of the system; meanwhile, its modular
  framework and open source R code are designed to facilitate expansion
  and problem-specific applications. We see high utility for this modeling
  framework in generating testable hypotheses and predicting integrated,
  mechanistic responses of corals to environmental change.

  Latest development--sigmaCH and sigmaCS changed to 0.1 meaning 10\% of
  metabolic CO2 produced is passively available to photosynthesis, to
  reflect the fact that that \textasciitilde{}10\% of total internal DIC
  at pH 7 is in the form of CO2. probably actually less than this since
  internal pH may be closer to 7.5. This does not change any of the
  figures substantively except the hysteresis figure--overshoot no longer
  occurs with specific parameters run. may need to just change light range
  or rate of change to reproduce overshoot.
  \end{abstract}

@@

The nutritional exchange between corals and \emph{Symbiodinium} directly
underlies the capacity of corals to build coral reef ecosystems, worth
trillions of US Dollars annually (Costanza, Groot, and Sutton 2014).
However, the complex symbiotic metabolism of corals is vulnerable to
disruption by numerous anthropogenic environmental perturbations,
jeopardizing their future persistence. In order to understand and
predict coral responses to complex changes in their environment, a
mechanistic understanding of how multiple interacting factors drive the
individual and emergent physiology of both symbiotic partners is
necessary. Such a task is well suited for theoretical modeling
frameworks such as Dynamic Energy Budget (DEB) theory (Kooijman 2010),
although the complexity of such theory makes these efforts inaccessible
to many biologists (Jager, Martin, and Zimmer 2013). In order to bridge
this gap, we present here a simplified dynamic bioenergetic model for
coral-\emph{Symbiodinium} symbioses that aims to mechanistically
integrate the impacts of complex environmental change on the
physiological performance of reef corals, including responses to
environmental stress.

@@

\begin{enumerate}
\def\labelenumi{\arabic{enumi}.}
\item
  \emph{Develop a detailed module of environmental stress.} Of primary
  interest to coral biologists and ecologists is symbiosis dysfunction
  under environmental stress, resulting in coral ``bleaching''--the loss
  of algal symbionts from the association (Jokiel and Coles 1977).
  Photooxidative stress in \emph{Symbiodinium} is considered a primary
  trigger of bleaching in response to high temperature and/or light
  (Weis 2008), and prolonged or severe bleaching can result in
  mortality, though corals sometimes recover their symbionts. Bleaching
  susceptibility, severity, and recovery may by influenced by
  interacting factors such as heterotrophy and nutrient availability
  (Wooldridge 2014b), and the genetic identity of \emph{Symbiodinium}
  (Glynn et al. 2001). To simulate these bleaching-related phenomena, we
  develop a generalized framework linking overreduction of the
  photosynthetic light reactions to downstream impacts of
  photoinhibition and photodamage.
\item
  \emph{Reduce theoretical and mathematical complexity.} Following the
  logic of Jager, Martin, and Zimmer (2013), we exclude certain features
  of formal DEB models in order to capture behaviors of interest with
  the simplest possible formulation. Here, we present a model without
  reserves, maturity, or reproduction (see Kooijman 2010). This
  formulation restricts the model's scope to the bioenergetics of growth
  and symbiosis dynamics in adult corals (i.e., reproduction, larval
  stages, and metamorphosis are not considered), but greatly reduces
  theoretical complexity and parameter numbers, which is advantageous
  given the relative paucity of data for corals. However, our primary
  motivation for reducing complexity was to increase accessibility and
  applicability for biologists and ecologists without requiring
  significant expertise in DEB theory.
\item
  \emph{Provide well-documented, open-access code.} In order to
  facilitate the continued development and application of theoretical
  modeling tools for coral symbioses, we provide open access to the
  model in the form of detailed, commented code written in the R
  language (R Core Team 2014). With an accessible and modular framework,
  we envision this model as a resource for futher development by the
  scientific community to include additional complexity and
  problem-specific components. To widen the audience for this work, we
  chose the R language because it is freely available and in common use
  by biologists and ecologists.
\end{enumerate}

With these as our primary motivations, we describe a simplified approach
to dynamic bioenergetic modeling of coral-algal symbioses that tracks
carbon and nitrogen acquisition and sharing between partners. This
theoretical framework dynamically integrates the influence of external
irradiance, nutrients, and prey availability on coral growth and
symbiosis dynamics (i.e., symbiont:host biomass ratios), allowing for
the possibility of coral bleaching in the event of photooxidative
stress. In the following sections, we describe the formulation of this
model and justify its structure and parameter values based on relevant
literature. We then demonstrate the model's behavior and discuss some of
its major implications and outcomes, and the wide range of potential
applications for this model in the study of cnidarian-algal symbioses.

@@

In this dynamic bioenergetic model of coral-algal symbiosis, carbon and
nitrogen are acquired by each partner and used to construct biomass. A
graphical representation of the model is presented in Fig. 1, and each
model flux and parameter are defined in Tables 1 and 2, respectively. We
use C-moles as the unit of biomass for consistency with the rigorous
mass balance of DEB theory: 1 C-mole is equivalent to the amount of
biomass containing 1 mole of Carbon atoms. Host biomass (\(H\)),
symbiont biomass (\(S\)), and prey biomass (\(X\)) have fixed, but
different, molar N:C ratios (Table 2). Carbon and nitrogen are combined
to produce biomass by synthesizing units (SU), which are mathematical
specifications of the formation of a product from two substrates; we use
the parallel complementary formulation of Kooijman (2010) to specify
these fluxes. The two state variables are symbiont biomass and coral
biomass; because resources are acquired proportionally to surface area,
and surface area is proportional to volume (i.e., corals are
``V1-morphs'' in DEB terminology (Kooijman 2010)), biomass increases
exponentially during growth (indeed, corals grow exponentially (Bak
1976)).

Environmental stress is implemented in the form of photo-oxidative
stress, which is thought to be the primary trigger of coral bleaching
(Lesser 1997; Weis 2008; Wooldridge 2009). To simulate these events, we
model the absorption and quenching of light energy by photochemistry and
non-photochemical quenching, and the phenomenological responses that
occur when these capacities are overwhelmed (i.e., photoinhibition,
photodamage, and symbiont loss). While bleaching in response to high
light alone has been observed experimentally (Schutter et al. 2011;
Downs et al. 2013), mass coral bleaching events occur in response to
high temperature (Hoegh-Guldberg 1999); thus, it is important to justify
our consideration of light as the primary stressor in the model. In
reality, light and temperature interact synergistically (Coles and
Jokiel 1978; Jones et al. 1998), and in fact, any stressor that disrupts
the quenching of light energy may lead to bleaching (Wooldridge 2010;
Baker and Cunning 2015). This is because the proximate cause of
photo-oxidative stress is excess excitation energy, but the upstream
events that lead to this situation may be diverse: indeed, elevated
temperature may inhibit Rubisco functioning (Jones et al. 1998) and the
repair of the D1 protein in photosystem II (Warner, Fitt, and Schmidt
1999), which reduces the capacity of photochemical quenching and leads
to an excess of light energy. In this way, elevated temperature serves
to reduce the threshold above which light stresses the system
(Hoegh-Guldberg 1999); importantly, light is still the proximate
stressor. Therefore, we omitted temperature from the model to maintain a
desired level of simplicity, while still allowing photophysiological
stress and bleaching to be simulated with biological realism in response
to light.

@@

\includegraphics{../img/Fig1.png}

@@

The balance equations for symbiont and host biomass are expressed as:

@@

where \(\rho_C\) is the surplus fixed carbon shared by the symbiont (see
below), and \(r_{NH}\) is the recycled portion of the nitrogen liberated
by host biomass turnover:

@@

This flux, \(j_{eC}\), is assumed to be available as a respiratory
substrate to support various energetically-demanding processes; of
particular importance is the host's active supply of CO\textsubscript{2}
for symbiont photosynthesis (Wooldridge 2013; Hopkinson, Tansik, and
Fitt 2015) (see below).

Due to the inherent inefficiency of the parallel complementary SU
formulation, there is always some nitrogen shared with the symbiont even
when coral biomass formation is strongly nitrogen-limited. Likewise,
there is always a non-zero rejection flux of excess carbon from the
coral biomass SU.

@@

We then specify the input of CO\textsubscript{2} to photosynthesis as
\(j_{CO_2}\), a flux mediated by the host that encompasses potentially
diverse carbon-concentrating mechanisms (CCMs), including active
transport of bicarbonate, carbonic anhydrase-catalyzed conversion of
bicarbonate to CO\textsubscript{2} to promote diffusion toward the
symbiont (Tansik, Fitt, and Hopkinson 2015), and acidification of the
symbiosome to increase localized CO\textsubscript{2} concentrations
around the symbiont (Barott et al. 2014). Since these active CCMs
require energetic input by the host, the magnitude of
CO\textsubscript{2} delivery is proportional to the amount of fixed
carbon available to the host after satisfying its growth demands,
\(j_{eC}\), assuming that some of this carbon is respired to energize
these CCMs. This formulation means that the symbiont indirectly ensures
its own CO\textsubscript{2} supply by providing fixed carbon (=energy)
to the host (Wooldridge 2013), which establishes a positive feedback in
the model that leads to the existence of multiple stable states, which
we discuss below in the context of coral bleaching. The parameter
\(k_{CO_2}\) scales the efficacy of host CCMs, which enables the
comparison of different rates of CO\textsubscript{2} delivery that may
characterize different coral species (Wooldridge 2014a). The active
input of CO\textsubscript{2} to the photosynthesis SU is therefore
specified as:

An additional fixed proportion of metabolic CO\textsubscript{2}
generated by host biomass turnover is passively made available to the
photosynthesis SU, according to:

@@

along with a fixed proportion of symbiont biomass turnover:

@@

where \(j_{CPm}\) is the maximum specific rate of photosynthesis, and
\(c_{ROS}\) is the relative rate of reactive oxygen species production
(see below). Dividing the photosynthetic rate by \(c_{ROS}\) causes a
decline in response to photooxidative stress at high light levels, and
the emergent outcome of this SU formulation demonstrates a classic
photoinhibition response (see Fig. S1).

Light energy absorbed in excess of what is used to fix carbon is
specified by the SU ``rejection flux'', according to:

@@

This excess light energy must be quenched by alternative pathways in
order to prevent photooxidative damage (Powles 1984).
\emph{Symbiodinium} utilize a variety of pathways for non-photochemical
quenching (NPQ; Roth 2014), which we collect in a relative NPQ capacity
specified as a parameter of the symbiont (\(k_{NPQ}\)). The total NPQ
flux is specified as a single-substrate SU formula:

@@

If light energy further exceeds the capacity of both photochemistry and
NPQ, then reactive oxygen species (ROS) are produced. We represent this
as a relative quantity \(c_{ROS}\), taking a value of 1 when all light
energy is quenched by photochemistry and NPQ, and increasing as the
amount of excess excitation energy increases:

@@

where \(k_{ROS}\) is a parameter of the symbiont that determines the
rate of ROS production (specifically, the amount of excess excitation
energy that doubles ROS production relative to baseline levels).
Importantly, \(c_{ROS}\) is specified here not as a function of absolute
external light, but rather the amount of excess light, \(j_{eL}\), after
accounting for quenching by carbon fixation and NPQ. A direct
consequence of this formulation is that CO\textsubscript{2}-limitation
of photosynthesis can lead to ROS production, an important mechanism
(Wooldridge 2009) that was not captured by previous representations of
photooxidative stress (Eynaud, Nisbet, and Muller 2011).

Carbon fixed by photosynthesis (\(j_{CP}\)) is then combined with
nitrogen shared by the host (\(\rho_N\)) and nitrogen recycled from
symbiont biomass turnover

@@

This second component of symbiont biomass loss represents both
photodamage and/or symbiont expulsion (i.e., bleaching), both of which
occur in response to high levels of ROS production. The parameter \(b\)
is included to scale biomass loss due to bleaching in response to ROS.
(Note that recycling of symbiont biomass turnover (\(rNS\) and \(rCS\))
only occurs based on the maintenance component of turnover (i.e.,
\(j_{ST}^0\)), and not the bleaching-related component, as this loss
represents biomass that is damaged or expelled from the host.)

@@

A time-stepping Euler method was used to solve the state equations since
the production and rejection fluxes of the SUs are imiplicitly defined
due to feedbacks present in the model. Specifically, the rejection
fluxes of carbon and nitrogen from the symbiont and host biomass SUs act
as reciprocal input fluxes to the other SU. In addition, the
photosynthesis SU receives CO\textsubscript{2} proportional to the
carbon rejection flux from the host biomass SU, and the rejection flux
of excitation energy from the photosynthesis SU acts to reduce its own
production (i.e., photoinhibition). A vector of time steps is used for
each simulation, along which dynamic environmental forcing functions
(irradiance, DIN, and prey abundance) can be designed. These vectors,
along with initial values of symbiont and host biomass, then serve as
input to the time-stepping function, which solves for the current system
state using values of the previous system state where necessary. A time
step of 0.1 days was used for all simulations, which were performed
using R code that is available in the data repository accompanying this
article: \url{github.com/jrcunning/Rcoral}.

@@

In a constant environment, the system ultimately reaches a steady state
of exponential growth or decline. However, under some conditions, either
of these outcomes may occur depending on initial values of symbiont and
host biomass, indicating the presence of alternate stable states. The
mechanism that produces these alternate stable states is the positive
feedback between carbon-limitation of the host and
CO\textsubscript{2}-limitation of photosynthesis: if symbiont biomass is
initially very low (i.e., a ``bleached'' coral), the system cannot
escape positively reinforcing carbon-limitation and cannot grow.
However, if symbiont biomass is initially high (i.e., a ``healthy''
coral), then the system maintains positive growth in a nitrogen-limited
state. For practical purposes, this section of the manuscript considers
only positive growth steady states under constant environments; the
subsequent ``dynamic behavior'' section explores how changing
environments may cause the system to switch between alternate stable
states, which we interpret in the context of coral bleaching.

>>>>

\usepackage[T1]{fontenc}
\usepackage{lmodern}
\usepackage{amssymb,amsmath}
\usepackage{ifxetex,ifluatex}
\usepackage{fixltx2e} % provides \textsubscript
% use upquote if available, for straight quotes in verbatim environments
\IfFileExists{upquote.sty}{\usepackage{upquote}}{}
\ifnum 0\ifxetex 1\fi\ifluatex 1\fi=0 % if pdftex
  \usepackage[utf8]{inputenc}
\else % if luatex or xelatex
  \usepackage{fontspec}
  \ifxetex
    \usepackage{xltxtra,xunicode}
  \fi
  \defaultfontfeatures{Mapping=tex-text,Scale=MatchLowercase}
  \newcommand{\euro}{€}
\fi
% use microtype if available
\IfFileExists{microtype.sty}{\usepackage{microtype}}{}
\usepackage{longtable}
\usepackage{graphicx}
% We will generate all images so they have a width \maxwidth. This means
% that they will get their normal width if they fit onto the page, but
% are scaled down if they would overflow the margins.
\makeatletter
\def\maxwidth{\ifdim\Gin@nat@width>\linewidth\linewidth
\else\Gin@nat@width\fi}
\makeatother
\let\Oldincludegraphics\includegraphics
\renewcommand{\includegraphics}[1]{\Oldincludegraphics[width=\maxwidth]{#1}}
\ifxetex
  \usepackage[setpagesize=false, % page size defined by xetex
              unicode=false, % unicode breaks when used with xetex
              xetex]{hyperref}
\else
  \usepackage[unicode=true]{hyperref}
\fi
\hypersetup{breaklinks=true,
            bookmarks=true,
            pdfauthor={},
            pdftitle={A simplified dynamic bioenergetic model for coral-Symbiodinium symbioses and coral bleaching as an alternate stable state},
            colorlinks=true,
            urlcolor=blue,
            linkcolor=magenta,
            pdfborder={0 0 0}}
\urlstyle{same}  % don't use monospace font for urls
\setlength{\parindent}{0pt}
\setlength{\parskip}{6pt plus 2pt minus 1pt}
\setlength{\emergencystretch}{3em}  % prevent overfull lines
\setcounter{secnumdepth}{0}
% Pandoc toggle for numbering sections (defaults to be off)
\setcounter{secnumdepth}{0}
% Pandoc header

@@

  \title{A simplified dynamic bioenergetic model for coral-\emph{Symbiodinium}
symbioses and coral bleaching as an alternate stable state}
    \author[University of Hawaii]{Ross Cunning\corref{c1}}
   \ead{ross.cunning@gmail.com} 
   \cortext[c1]{Corresponding Author}
    \author[University of California]{Erik B. Muller}

@@

  \begin{abstract}
  Coral reef ecosystems owe their ecological success--and vulnerability to
  climate change--to the symbiotic metabolism of corals and
  \emph{Symbiodinium} spp. The urgency to understand and predict the
  stability and breakdown of these symbioses (i.e., coral `bleaching')
  demands the development and application of theoretical tools; however,
  the complexity of some approaches may limit their potential application
  and utility. Here, we develop a simplified dynamic bioenergetic model of
  coral-\emph{Symbiodinium} symbioses that demonstrates realistic
  steady-state patterns in coral growth and symbiont abundance across
  gradients of light, nutrients, and feeding. Furthermore, through a
  mechanistic treatment of photo-oxidative stress, the model displays
  dynamic behaviors of bleaching and recovery, and suggests that these
  phenomena represent transitions between alternate stable states. A suite
  of complex responses to multiple, interacting environmental factors
  reproduced by the model suggests it unifyingly captures many important
  attributes of the system; meanwhile, its modular framework and open
  source R code are designed to facilitate expansion and problem-specific
  applications. We see significant potential applications for this
  modeling framework in generating testable hypotheses and predicting
  integrated, mechanistic responses of corals to environmental change.
  \end{abstract}

@@

The nutritional exchange between corals and \emph{Symbiodinium} directly
underlies the capacity of corals to build coral reef ecosystems, worth
trillions of US Dollars annually (Costanza, Groot, and Sutton 2014).
However, the complex symbiotic metabolism of corals is vulnerable to
disruption by numerous anthropogenic environmental perturbations,
jeopardizing their future persistence. In order to understand and
predict responses of corals to complex changes in their environment, a
mechanistic understanding of how multiple interacting factors drive the
individual and emergent physiology of both symbiotic partners is
necessary. Such a task is well suited for theoretical modeling
frameworks such as Dynamic Energy Budget (DEB) theory (Kooijman 2010),
although the complexity of such theory makes these efforts inaccessible
to many biologists (Jager, Martin, and Zimmer 2013). In order to bridge
this gap, we present here a simplified dynamic bioenergetic model for
coral-\emph{Symbiodinium} symbioses that aims to mechanistically
integrate the impacts of complex environmental change on the
physiological performance of reef corals, including responses to
environmental stress.

@@

\begin{enumerate}
\def\labelenumi{\arabic{enumi}.}
\item
  \emph{Develop a module of photooxidative stress.} Of primary interest
  to coral biologists and ecologists is symbiosis dysfunction under
  environmental stress, resulting in coral ``bleaching''--the loss of
  algal symbionts from the association (Jokiel and Coles 1977).
  Photooxidative stress in \emph{Symbiodinium} is considered a primary
  trigger of bleaching in response to high temperature and/or light
  (Weis 2008), and prolonged or severe bleaching can result in
  mortality, though corals sometimes recover their symbionts. Bleaching
  susceptibility, severity, and recovery may be influenced by
  interacting factors such as heterotrophy and nutrient availability
  (Wooldridge 2014b). To simulate these events and interactions, we
  develop a photooxidative stress module linking overreduction of the
  photosynthetic light reactions to downstream impacts of
  photoinhibition and photodamage.
\item
  \emph{Reduce theoretical and mathematical complexity.} Following the
  logic of Jager, Martin, and Zimmer (2013), we exclude certain features
  of formal DEB theory in order to capture behaviors of interest with
  the simplest possible formulation. Here, we present a model without
  reserves, maturity, or reproduction (see Kooijman 2010). This
  formulation restricts the model's scope to the bioenergetics of adult
  corals (i.e., reproduction, larval stages, and metamorphosis are not
  considered), but greatly reduces theoretical complexity and parameter
  numbers, which is advantageous given the relative paucity of data for
  corals. However, our primary motivation for reducing complexity was to
  increase accessibility and applicability for biologists and ecologists
  without requiring significant expertise in DEB theory.
\item
  \emph{Provide well-documented, open-access code.} In order to
  facilitate the continued development and application of theoretical
  modeling tools for coral symbioses, we provide open access to the
  model in the form of detailed, commented code written in the R
  language (R Core Team 2014). With an accessible and modular framework,
  we envision this model as a resource for futher development by the
  scientific community to include additional complexity and
  problem-specific components. We chose the R language, which is freely
  available and in common use by biologists and ecologists, to reach the
  widest possible audience with this work.
\end{enumerate}

With these as our primary motivations, we describe a simplified approach
to dynamic bioenergetic modeling of coral-\emph{Symbiodinium} symbioses
that dynamically integrates the influence of external irradiance,
nutrients, and prey availability on coral growth and symbiosis dynamics
(i.e., symbiont:host biomass ratios), allowing for the possibility of
coral bleaching in the event of photooxidative stress. In the following
sections, we describe and provide rationale for the model structure, and
demonstrate a range of steady state and dynamic behaviors that are
consistent with observed phenomena.

@@

In this dynamical system, both partners acquire and use carbon and
nitrogen to construct biomass: the symbiont fixes carbon through
photosynthesis, and receives nitrogen shared by the host, while the host
acquires nitrogen from the environment and receives carbon shared by the
symbiont. A graphical representation of the model is presented in Fig.
1, and each model flux and parameter is defined in Tables 1 and 2,
respectively. We use C-moles as the unit of biomass for consistency with
the rigorous mass balance of DEB theory: 1 C-mole is equivalent to the
amount of biomass containing 1 mole of carbon atoms. Host biomass
(\(H\)), symbiont biomass (\(S\)), and prey biomass (\(X\)) have fixed,
but different, molar N:C ratios (Table 2). Biomass is produced from
carbon and nitrogen by synthesizing units (SU), which are mathematical
specifications of the formation of a product from two substrates; we use
the parallel complementary formulation of Kooijman (2010) to specify
these fluxes. The two state variables of this system are symbiont
biomass and coral biomass; because resources are acquired proportionally
to surface area, and surface area is proportional to volume (i.e.,
corals are ``V1-morphs'' in DEB terminology (Kooijman 2010)), biomass
increases exponentially during growth (indeed, corals grow exponentially
(Bak 1976)).

Environmental stress is implemented in the form of photooxidative
stress, which is thought to be the primary trigger of coral bleaching
(Lesser 1997; Weis 2008; Wooldridge 2009). To simulate these events, we
model the absorption and quenching of light energy by photochemistry and
non-photochemical quenching, and the phenomenological responses that
occur when these capacities are overwhelmed (i.e., photoinhibition,
photodamage, and symbiont loss). While bleaching in response to high
light alone has been observed experimentally (Schutter et al. 2011;
Downs et al. 2013), mass coral bleaching events occur in response to
high temperature (Hoegh-Guldberg 1999); thus, it is important to justify
our consideration of light as the primary stressor. In reality, light
and temperature interact synergistically (Coles and Jokiel 1978; Jones
et al. 1998), and in fact, any stressor that disrupts the quenching of
light energy may lead to bleaching (Wooldridge 2010; Baker and Cunning
2015). This is because the proximate cause of photo-oxidative stress is
excess excitation energy, but the upstream events that lead to this
situation may be diverse: indeed, elevated temperature may inhibit
Rubisco functioning (Jones et al. 1998) and the repair of the D1 protein
in photosystem II (Warner, Fitt, and Schmidt 1999), which reduces the
capacity of photochemical quenching and leads to an excess of light
energy. In this way, elevated temperature serves to reduce the threshold
above which light stresses the system (Hoegh-Guldberg 1999);
importantly, light is still the proximate stressor. Therefore, we
omitted temperature from the model to maintain a desired level of
simplicity, while still allowing photooxidative stress and bleaching to
be simulated with biological realism in response to light.

@@

\begin{figure}[htbp]
\centering
\includegraphics{../img/Fig1.png}
\caption{Graphical representation of coral-algal symbiosis model. Light,
CO\textsubscript{2}, prey, and DIN are acquired from the external
environment proportional to the biomass of the partner indicated by the
black box for uptake. Mass fluxes (see Table 1 for definitions) are
represented by \(j\)'s with subscripts indicating the type of mass, and
in some cases the process (e.g., \(j_{CP}\) is the flux of carbon
produced by photosynthesis), and \(\rho\)'s indicate fluxes that are
shared by one partner with the other. Parallel complementary
synthesizing units (SUs) are represented by large circles, and rejection
fluxes from these SUs are indicated by dashed lines. \(c_{ROS}\) is a
proportional rate that impacts other model fluxes by inhibition or
acceleration; likewise, \(j_{eC}\) accelerates the rate of \(j_{CO_2}\).
Recycled mass fluxes from biomass turnover are not shown for clarity
(but see Table 1 for definitions).}
\end{figure}

@@

The balance equations for symbiont (\(S\)) and host (\(H\)) biomass are
expressed as:

@@

where \(\rho_C\) is fixed carbon shared by the symbiont (see Eq. XX),
and \(r_{NH}\) is recycled nitrogen liberated by host biomass turnover:

@@

This flux, \(j_{eC}\), is assumed to be available to the host as a
respiratory substrate to support various energetically-demanding
processes; of particular importance is the host's active carbon
concentrating mechanisms (CCMs) that supply CO\textsubscript{2} for
symbiont photosynthesis (Wooldridge 2013; Hopkinson, Tansik, and Fitt
2015). We therefore specify \(j_{CO_2}\) as the host-mediated delivery
of CO\textsubscript{2} to photosynthesis that encompasses potentially
diverse carbon-concentrating mechanisms (CCMs), including active
transport of bicarbonate, carbonic anhydrase-catalyzed conversion of
bicarbonate to CO\textsubscript{2} to promote diffusion toward the
symbiont (Tansik, Fitt, and Hopkinson 2015), and acidification of the
symbiosome to increase localized CO\textsubscript{2} concentrations
around the symbiont (Barott et al. 2014). Since these active CCMs
require energetic input by the host, we define \(j_{CO_2}\) as
proportional to \(j_{eC}\), assuming that some of this carbon is
respired to energize the CCMs. This formulation means that the symbiont
indirectly ensures its own CO\textsubscript{2} supply by providing fixed
carbon (=energy) to the host (Wooldridge 2013), which establishes a
positive feedback in the model that leads to the existence of multiple
stable states, which we discuss below in the context of coral bleaching.
The parameter \(k_{CO_2}\) scales the efficacy of host CCMs, which
enables the comparison of different rates of CO\textsubscript{2}
delivery that may characterize different coral species (Wooldridge
2014a). The active input of CO\textsubscript{2} to the photosynthesis SU
is therefore specified as:

@@

CO\textsubscript{2} arrives at the photosynthesis SU from multiple
sources: in addition to the CO\textsubscript{2} actively supplied by the
host through its CCMs (\(j_{CO_2}\); Eq. XX), a fixed proportion of
metabolic CO\textsubscript{2} generated by host biomass turnover is
passively available to the photosynthesis SU, according to:

@@

along with a fixed proportion of CO\textsubscript{2} generated by
symbiont biomass turnover:

@@

where \(j_{CPm}\) is the maximum specific rate of photosynthesis, and
\(c_{ROS}\) is the relative rate of reactive oxygen species production
(see Eq. XX). Dividing the photosynthetic rate by \(c_{ROS}\) causes a
decline in response to photooxidative stress at high light levels, and
the emergent outcome of this SU formulation demonstrates a classic
photoinhibition response (Fig. S2).

Light energy absorbed in excess of what is used to fix carbon is
specified by the SU rejection flux, according to:

@@

This excess light energy must be quenched by alternative pathways in
order to prevent photooxidative damage (Powles 1984).
\emph{Symbiodinium} utilize a variety of pathways for non-photochemical
quenching (NPQ; Roth 2014), which we collect in a total NPQ capacity
specified as a parameter of the symbiont (\(k_{NPQ}\)). The NPQ flux
\(j_{NPQ}\) is then specified as a single-substrate SU formula with a
maximum of \(k_{NPQ}\):

@@

If light energy further exceeds the capacity of both photochemistry and
NPQ, then reactive oxygen species (ROS) are produced. We represent this
as a relative quantity \(c_{ROS}\), which takes a value of 1 when all
light energy is quenched by photochemistry and NPQ, and increases as the
amount of excess excitation energy increases, specified as:

@@

where \(k_{ROS}\) is a parameter of the symbiont that determines the
rate of ROS production (specifically, the amount of excess excitation
energy that doubles ROS production relative to baseline levels).
Importantly, \(c_{ROS}\) is specified here not as a function of absolute
external light, but rather the amount of excess light after accounting
for quenching by carbon fixation and NPQ. A direct consequence of this
formulation is that CO\textsubscript{2}-limitation of photosynthesis can
lead to ROS production, an important mechanism (Wooldridge 2009) that
was not captured by previous representations of photooxidative stress
(Eynaud, Nisbet, and Muller 2011).

Carbon fixed by photosynthesis (\(j_{CP}\); Eq. XX) is then combined
with nitrogen shared by the host (\(\rho_N\); Eq. XX) and nitrogen
recycled from symbiont biomass turnover

@@

This second component of symbiont biomass loss represents both
photodamage and/or symbiont expulsion (i.e., bleaching), both of which
occur in response to high levels of ROS production. The parameter \(b\)
is included to scale biomass loss due to bleaching in response to ROS.
(Note that recycling of symbiont biomass turnover (\(rNS\) and \(rCS\))
only occurs based on the maintenance component of turnover (i.e.,
\(j_{ST}^0\)), and not the photodamage/bleaching component, as this loss
represents biomass that is damaged or expelled from the host.)

@@

A time-stepping Euler method was used to solve the state equations since
the production and rejection fluxes of the SUs are imiplicitly defined
due to feedbacks present in the model. Specifically, the rejection
fluxes of carbon and nitrogen from the symbiont and host biomass SUs act
as reciprocal input fluxes to the other SU. In addition, the
photosynthesis SU receives CO\textsubscript{2} proportional to the
carbon rejection flux from the host biomass SU, and the rejection flux
of excitation energy from the photosynthesis SU acts to reduce its own
production (i.e., photoinhibition). A vector of time steps is used for
each simulation, along which dynamic environmental forcing functions
(irradiance, DIN, and prey abundance). These vectors, along with initial
values of symbiont and host biomass, then serve as input to the
time-stepping function, which solves for the current system state using
values of the previous system state where necessary. A time step of 0.1
days was used for all simulations, which were performed using R code
that is available in the data repository accompanying this article:
\url{github.com/jrcunning/Rcoral}. For steady state estimations,
simulations were run until the change in both growth and S:H biomass was
\textless{} 5e-7 per time step.

@@

In a constant environment, the system ultimately reaches a steady state
of exponential growth or decline. However, under some conditions, either
of these outcomes may occur depending on initial values of symbiont and
host biomass, indicating the presence of alternate stable states. The
mechanism that produces these alternate stable states is the positive
feedback between carbon-limitation of the host and
CO\textsubscript{2}-limitation of photosynthesis: if symbiont biomass is
initially very low (i.e., a ``bleached'' coral), the system cannot
escape this positive feedback and cannot grow. However, if symbiont
biomass is initially high (i.e., a ``healthy'' coral), then the system
remains in a nitrogen-limited state with positive growth. For practical
purposes, this section of the manuscript considers only positive growth
steady states under constant environments; the subsequent ``dynamic
behavior'' section explores how changing environments may cause the
system to switch between alternate stable states, which we interpret in
the context of coral bleaching.

To analyze steady state behavior, we ran the model to steady state
across gradients of external irradiance and nutrients (Fig. 2), which
revealed patterns consistent with observed phenomena in corals.
Predicted growth rates are low at low light and DIN
(\textasciitilde{}0.01 d\textsuperscript{-1}), and begin increasing as
both of these factors increase (Fig. 2A). Low light limits
photosynthetic rates, resulting in less fixed carbon shared with the
host and an associated increase in the symbiont to host biomass ratio
(Fig. 2B). In agreement with this trend are many observations of
negative correlation between irradiance and symbiont density (Stimson
1997; Brown et al. 1999; Fitt et al. 2000; Titlyanov et al. 2001). As
higher light alleviates light-limitation of photosynthesis, host growth
becomes less carbon-limited.
